Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-26 paper-new/categorical-model.tex
changes to categorical model

Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
distinguished morphism $\delta_A^*$, such that $\delta_A^* \bisim_{FA, FA} \id_{FA}$.

@@

\begin{definition}\label{def:step-1-model}
A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
with the additional categories and functors for bisimiarity described above, as  well as
the existence of a distinguised computation morphism $\delta_A^* \bisim \id_{FA}$ for each $A$.
\end{definition}

@@

We require that $\delta_A^* \in \pe_{FA}$ for all $A$, where $\delta_A^*$ is the distinguished
morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
of a step-1 model.

@@

\begin{definition}\label{def:step-2-model}
  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
  \begin{enumerate}
    \item For each value type $A$, there is a monoid $\pv_A$ that is a submonoid of
    $\{ f \in \vf(A,A) \mid f \bisim \id \}$.
    \item For each computation type $B$, there is a monoid $\pe_B$ that is a submonoid of
    $\{ g \in \ef(B,B) \mid g \bisim \id \}$.
    \item For all $A$, the distinguished endomorphism $\delta_A^*$ is in $\pe_{FA}$.
    % \item $\pv_A$ and a monoid homomorphism 
    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
    % \item $\pe_B$ and a monoid homomorphism
    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
  \end{enumerate}
\end{definition}

@@

As is the case in the extensional model, there is a relationship between
vertical (i.e., function) morphisms and horizontal (i.e., relation) morphisms,
but as mentioned above, now there
are perturbations involved in order to keep both sides ``in lock-step".
The precise definitions are as follows.

% TODO: Give these squares names
\begin{definition}\label{def:quasi-left-representable}
Let $c : A \rel A'$ be a value relation. We say that $c$ is \emph{quasi-left-representable by}
$f \in \vf(A, A')$ if there are perturbations $\delta_c^{l,e} \in \pv_A$ and
$\delta_c^{r,e} \in \pv_{A'}$ such that the following squares commute:

\begin{center}
  \begin{tabular}{ m{7em} m{7em} } 
    % UpL
    \begin{tikzcd}[ampersand replacement=\&]
      A \& {A'} \\
      {A'} \& {A'}
      \arrow["f"', from=1-1, to=2-1]
      \arrow["\delta_c^{r,e}", from=1-2, to=2-2]
      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow[from=2-1, to=2-2, Rightarrow, no head]
    \end{tikzcd}
    &
    % UpR
    \begin{tikzcd}[ampersand replacement=\&]
      A \& {A} \\
      {A} \& {A'}
      \arrow["\delta_c^{l,e}"', from=1-1, to=2-1]
      \arrow["f", from=1-2, to=2-2]
      \arrow[from=1-1, to=1-2, Rightarrow, no head]
      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
    \end{tikzcd}
  \end{tabular}
\end{center}

@@

\begin{definition}\label{def:quasi-right-representable}
Let $d : B \rel B'$ be a computation relation. We say that $d$ is
\emph{quasi-right-representable by} $f \in \ef(B', B)$
if there exist perturbations $\delta_d^{l,p} \in \pe_B$ and
$\delta_d^{r,p} \in \pe_{B'}$ such that the following squares commute:

\begin{center}
  \begin{tabular}{ m{7em} m{7em} } 
    % DnR
    \begin{tikzcd}[ampersand replacement=\&]
      {B} \& {B'} \\
      {B} \& {B}
      \arrow["\delta_d^{l,p}"', from=1-1, to=2-1]
      \arrow["g", from=1-2, to=2-2]
      \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow[from=2-1, to=2-2, Rightarrow, no head]
    \end{tikzcd}
    &
    % DnL
    \begin{tikzcd}[ampersand replacement=\&]
      {B'} \& {B'} \\
      {B} \& {B'}
      \arrow["g"', from=1-1, to=2-1]
      \arrow["\delta_d^{r,p}", from=1-2, to=2-2]
      \arrow[from=1-1, to=1-2, Rightarrow, no head]
      \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
    \end{tikzcd}
  \end{tabular}
\end{center}

@@

@@

Lastly, we require that the model satisfy a weak version of functoriality for 
the CBPV connectives $U,F,\times,\to$. 
First, we will need a definition:

@@

We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
which we specify as follows:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d)U(d')$
  \item $F(c \comp c'') \qordeq F(c)F(c')$
  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
\end{itemize}

@@

\begin{definition}\label{def:step-3-model}
  A \emph{step-3 intensional model} consists
  of all the data of a step-2 intensional model, such that additionally:
  \begin{enumerate}
    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
  \end{enumerate}
\end{definition}

% Want: U d \comp U d' = U(d \comp d')
%       F c \comp F c' = F(c \comp c')
% Add requirement: Either the model is functorial with respect to up/downcasts or with repsect to relations
% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsubsection{Adding Perturbations}

@@

We claim that from a step-1 model, we can construct a step-2 model. 
% TODO give overview
For the proof, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.

@@

% TODO give overview

>>>>

@@

% Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
% distinguished morphism $\delta_{FA}^*$, such that $\delta_A^* \bisim_{FA, FA} \id_{FA}$.
% Moreover, we require that these morphisms are related in that for any
% $c : A \rel A'$, we have a square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$.

Lastly, we require that for any computation object $B$, the hom-set $\vf(UB, UB)$ contains a
distinguished morphism $\delta_{UB}^*$, such that $\delta_{UB}^* \bisim_{UB, UB} \id_{UB}$.
Moreover, we require that these morphisms are related in that for any
$d : B \rel B'$, we have a square $\delta_{UB}^* \ltdyn_{Ud}^{Ud} \delta_{UB'}^*$.
We also require that these morphisms commute with computation morphisms, in the sense
that for any $\phi \in \ef(B, B')$ we have $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$.

Given the existence of the morphisms $\delta_{UB}^*$, we can define computation morphisms
$\delta_{FA}^* \in \ef(FA, FA)$ for all $A$ using the

\begin{definition}\label{def:step-1-model}
A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
with:
\begin{itemize}
  \item The categories and functors for bisimiarity described above.
  \item The existence of a distinguished value morphism $\delta_{UB}^* \bisim \id_{UB}$ for each $B$.
  \item A square $\delta_{UB}^* \ltdyn_{Ud}^{Ud} \delta_{UB'}^*$ for all $d : B \rel B'$.
  \item The commutativity condition $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$ for any $\phi \in \ef(B, B')$.
\end{itemize}
% We also require the existence of a distinguised computation morphism $\delta_{FA}^* \bisim \id_{FA}$ for each $A$,
% and a square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A'$.
\end{definition}

@@

We require that $\delta_{FA}^* \in \pe_{FA}$ for all $A$, where $\delta_{FA}^*$ is the distinguished
morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
of a step-1 model.

@@

\begin{definition}\label{def:step-2-model}
  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
  \begin{enumerate}
    \item For each value type $A$, there is a monoid $\pv_A$ that is a submonoid of
    $\{ f \in \vf(A,A) \mid f \bisim \id \}$.
    \item For each computation type $B$, there is a monoid $\pe_B$ that is a submonoid of
    $\{ g \in \ef(B,B) \mid g \bisim \id \}$.
    \item For all $A$, the distinguished endomorphism $\delta_{FA}^*$ is in $\pe_{FA}$.
    % \item $\pv_A$ and a monoid homomorphism 
    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
    % \item $\pe_B$ and a monoid homomorphism
    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
  \end{enumerate}
\end{definition}

@@

As is the case in the extensional model, there is a relationship between
vertical (i.e., function) morphisms and horizontal (i.e., relation) morphisms,
but as mentioned above, now there
are perturbations involved in order to keep both sides ``in lock-step".
We begin with a step-2 intensional model as defined in the previous section,
and provide the additional conditions that axiomatize the behavior of casts.
The precise definitions are as follows.

First, let $\mathcal M$ be any double category with a notion of perturbations,
i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ which is a
submonoid of the endomorphisms on $X$.

\begin{definition}\label{def:quasi-left-representable}
  Let $R$ be a horizontal morphism in $\mathcal M$ between objects $X$ and $Y$.
  We say that $R$ is quasi-left-representable by a vertical morphism $f$ in
  $\mathcal M(X, Y)$ if there are perturbations $\delle_R \in P_X$ and
  $\delre_R \in P_Y$ such that the following squares commute:

  \begin{center}
    \begin{tabular}{ m{7em} m{7em} } 
      % UpL
      \begin{tikzcd}[ampersand replacement=\&]
        X \& {Y} \\
        {Y} \& {Y}
        \arrow["f"', from=1-1, to=2-1]
        \arrow["\delta_R^{r,e}", from=1-2, to=2-2]
        \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow[from=2-1, to=2-2, Rightarrow, no head]
      \end{tikzcd}
      &
      % UpR
      \begin{tikzcd}[ampersand replacement=\&]
        X \& {X} \\
        {X} \& {Y}
        \arrow["\delta_R^{l,e}"', from=1-1, to=2-1]
        \arrow["f", from=1-2, to=2-2]
        \arrow[from=1-1, to=1-2, Rightarrow, no head]
        \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \end{tikzcd}
    \end{tabular}
  \end{center}

@@

\begin{definition}\label{def:quasi-right-representable}
  Let $R$ be a horizontal morphism between $X$ and $Y$. We say that $R$ is
  \emph{quasi-right-representable by} $f \in \mathcal M(Y, X)$
  if there exist perturbations $\dellp_R \in P_X$ and
  $\delrp_R \in P_Y$ such that the following squares commute:

  \begin{center}
    \begin{tabular}{ m{7em} m{7em} } 
      % DnR
      \begin{tikzcd}[ampersand replacement=\&]
        {X} \& {Y} \\
        {X} \& {X}
        \arrow["\delta_R^{l,p}"', from=1-1, to=2-1]
        \arrow["f", from=1-2, to=2-2]
        \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow[from=2-1, to=2-2, Rightarrow, no head]
      \end{tikzcd}
      &
      % DnL
      \begin{tikzcd}[ampersand replacement=\&]
        {Y} \& {Y} \\
        {X} \& {Y}
        \arrow["f"', from=1-1, to=2-1]
        \arrow["\delta_R^{r,p}", from=1-2, to=2-2]
        \arrow[from=1-1, to=1-2, Rightarrow, no head]
        \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \end{tikzcd}
    \end{tabular}
  \end{center}

@@

% TODO: Give these squares names
% \begin{definition}\label{def:quasi-left-representable}
% Let $c : A \rel A'$ be a value relation. We say that $c$ is \emph{quasi-left-representable by}
% $f \in \vf(A, A')$ if there are perturbations $\delta_c^{l,e} \in \pv_A$ and
% $\delta_c^{r,e} \in \pv_{A'}$ such that the following squares commute:

% \begin{center}
%   \begin{tabular}{ m{7em} m{7em} } 
%     % UpL
%     \begin{tikzcd}[ampersand replacement=\&]
%       A \& {A'} \\
%       {A'} \& {A'}
%       \arrow["f"', from=1-1, to=2-1]
%       \arrow["\delta_c^{r,e}", from=1-2, to=2-2]
%       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%       \arrow[from=2-1, to=2-2, Rightarrow, no head]
%     \end{tikzcd}
%     &
%     % UpR
%     \begin{tikzcd}[ampersand replacement=\&]
%       A \& {A} \\
%       {A} \& {A'}
%       \arrow["\delta_c^{l,e}"', from=1-1, to=2-1]
%       \arrow["f", from=1-2, to=2-2]
%       \arrow[from=1-1, to=1-2, Rightarrow, no head]
%       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%     \end{tikzcd}
%   \end{tabular}
% \end{center}

% We call the first square $\upl$ and the second square $\upr$.

% \end{definition}

% \begin{definition}\label{def:quasi-right-representable}
% Let $d : B \rel B'$ be a computation relation. We say that $d$ is
% \emph{quasi-right-representable by} $f \in \ef(B', B)$
% if there exist perturbations $\delta_d^{l,p} \in \pe_B$ and
% $\delta_d^{r,p} \in \pe_{B'}$ such that the following squares commute:

% \begin{center}
%   \begin{tabular}{ m{7em} m{7em} } 
%     % DnR
%     \begin{tikzcd}[ampersand replacement=\&]
%       {B} \& {B'} \\
%       {B} \& {B}
%       \arrow["\delta_d^{l,p}"', from=1-1, to=2-1]
%       \arrow["g", from=1-2, to=2-2]
%       \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%       \arrow[from=2-1, to=2-2, Rightarrow, no head]
%     \end{tikzcd}
%     &
%     % DnL
%     \begin{tikzcd}[ampersand replacement=\&]
%       {B'} \& {B'} \\
%       {B} \& {B'}
%       \arrow["g"', from=1-1, to=2-1]
%       \arrow["\delta_d^{r,p}", from=1-2, to=2-2]
%       \arrow[from=1-1, to=1-2, Rightarrow, no head]
%       \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%     \end{tikzcd}
%   \end{tabular}
% \end{center}

% We call the first square $\dnr$ and the second square $\dnl$.

% \end{definition}

With these definitions, we return to the more specific setting of a step-2
intensional model $\mathcal M$ and specify the new requirements for relations.
We require that there are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$
Every value edge $c : A \rel A'$ must be quasi-left-representable by $\upf(c)$,
and every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.

@@

\begin{comment}
Lastly, we require that the model satisfy a weak version of functoriality for 
the CBPV connectives $U,F,\times,\to$. 
First, we will need a definition:

@@

We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
which we specify as follows:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d)U(d')$
  \item $F(c \comp c'') \qordeq F(c)F(c')$
  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
\end{itemize}
\end{comment}

@@

\begin{definition}\label{def:step-3-model}
  A \emph{step-3 intensional model} consists
  of all the data of a step-2 intensional model, such that additionally:
  \begin{enumerate}
    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
    \item \eric{Do we need the quasi-functoriality of CBPV connectives on relations?}
    % \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
  \end{enumerate}
\end{definition}

% Want: U d \comp U d' = U(d \comp d')
%       F c \comp F c' = F(c \comp c')
% Add requirement: Either the model is functorial with respect to up/downcasts or with repsect to relations
%

@@

\subsubsection{Adding Perturbations}\label{sec:constructing-perturbations}

@@

We claim that from a step-1 model, we can construct a step-2 model. 
\eric{TODO give overview}
For the proof, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.

@@

\eric{TODO give overview}
